---
title: "Problem Set 4 — Credit Card Expenditure Analysis"
author: "Joachim Vogl"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Overview

This analysis uses the `CreditCard` dataset from the `AER` package to examine
how **age**, **number of dependents**, and **self-employment status** relate to
monthly credit card expenditure. All standard errors are heteroskedasticity-robust
(HC1) throughout.

```{r load-packages-data}
library(AER)
library(lmtest)
library(sandwich)

data(CreditCard)
```

---

## Part (a): Linear Regression with Robust Standard Errors

A multiple regression model is estimated with `expenditure` as the dependent
variable and `dependents`, `selfemp`, and `age` as predictors. Because the
dataset exhibits heteroskedasticity, standard OLS standard errors are
unreliable — the table below uses HC1 heteroskedasticity-robust standard errors.

```{r part-a-linear-model}
linear.model.binary <- lm(expenditure ~ dependents + selfemp + age,
                           data = CreditCard)

coeftest(linear.model.binary,
         vcov = vcovHC(linear.model.binary, type = "HC1"))
```

**Interpretation:**

- **Dependents (β = 11.49):** Holding self-employment status and age constant,
  each additional dependent is associated with approximately $11.49 more in
  average monthly credit card expenditure. This estimate is statistically
  significant at the 10% level (p = 0.080) but not at the conventional 5% level.

- **Self-employment (β = −41.47):** Holding other variables constant,
  self-employed individuals spend on average about $41.47 less per month than
  non-self-employed individuals. This difference is not statistically significant
  (p = 0.130).

- **Age (β = 0.21):** Each additional year of age is associated with only about
  $0.21 more in monthly expenditure — an economically negligible effect that is
  not statistically significant (p = 0.777).

**Summary:** `dependents` is the only statistically significant predictor, and
only at the 10% level. Neither `selfemp` nor `age` is significant at any
conventional level.

---

## Part (b): Joint F-Test — Dependents and Age

The null hypothesis is that the coefficients on `dependents` and `age` are
**jointly equal to zero**. A robust Wald (F) test is used.

```{r part-b-joint-ftest}
linearHypothesis(linear.model.binary,
                 c("dependents = 0", "age = 0"),
                 vcov. = vcovHC(linear.model.binary, type = "HC1"))
```

**Interpretation:** The F-statistic is 1.631 with a p-value of 0.196. Because
p > 0.05, we **fail to reject** the null hypothesis that `dependents` and `age`
are jointly zero. There is insufficient evidence that these two variables
together explain variation in expenditure.

---

## Part (c): Quadratic Model in Age

To allow for a nonlinear relationship between age and expenditure, a quadratic
term (`age²`) is added to the model.

```{r part-c-quadratic-model}
quadratic.model <- lm(expenditure ~ age + I(age^2) + dependents + selfemp,
                       data = CreditCard)

coeftest(quadratic.model,
         vcov = vcovHC(quadratic.model, type = "HC1"))
```

**Interpretation:** The positive coefficient on `age` (+5.15) and negative
coefficient on `age²` (−0.066) together indicate that expenditure increases with
age, but at a **decreasing rate** — a concave (inverted-U) relationship. Both
terms are statistically significant at the 5% level (age: p = 0.036;
age²: p = 0.033), confirming that the nonlinearity is genuine rather than
sampling noise.

---

## Part (d): Linear-Log Model

A linear-log specification replaces the raw `age` variable with its natural
log, allowing a percentage-change interpretation.

```{r part-d-linearlog-model}
linearlog.model <- lm(expenditure ~ log(age) + dependents + selfemp,
                       data = CreditCard)

coeftest(linearlog.model,
         vcov = vcovHC(linearlog.model, type = "HC1"))
```

**Interpretation:**

The estimated equation is:

> *Expenditure = 162.33 + 4.07 · ln(age) + 11.61 · dependents − 41.18 · selfemp*

- **log(age) (β = 4.07):** A 1% increase in age is associated with an increase
  of approximately $0.04 in monthly expenditure (4.07 ÷ 100 ≈ 0.041). This
  effect is economically negligible and not statistically significant (p = 0.783).

- **Dependents (β = 11.61):** Each additional dependent is associated with about
  $11.61 more in monthly expenditure. Statistically significant at the 10% level
  (p = 0.077).

- **Self-employment (β = −41.18):** Self-employed individuals spend about $41.18
  less per month than non-self-employed individuals. Not statistically significant
  (p = 0.128).

---

## Visualization: Linear vs. Quadratic Age Fit

The scatterplot below shows individual-level expenditure by age, overlaid with
the linear (green) and quadratic (red) fitted regression lines.

```{r plot-age-expenditure, fig.width=8, fig.height=5}
plot(CreditCard$age, CreditCard$expenditure,
     col  = "steelblue",
     pch  = 20,
     xlab = "Age (years)",
     ylab = "Monthly Expenditure ($)",
     main = "Estimated Linear and Quadratic Regression Functions")

lin_age  <- lm(expenditure ~ age, data = CreditCard)
abline(lin_age, col = "green", lwd = 2)

quad_age  <- lm(expenditure ~ age + I(age^2), data = CreditCard)
age_order <- order(CreditCard$age)
lines(x = CreditCard$age[age_order],
      y = fitted(quad_age)[age_order],
      col = "red",
      lwd = 2)

legend("topright",
       legend = c("Linear", "Quadratic"),
       col    = c("green", "red"),
       lwd    = 2)
```
